---
title: "Wine Chemistry and Correlates of Quality"
subtitle: "Stat 184 Final Project"
author: "Jason Kaufmann, Akshith Kachakayala, Joshua Tanyi"
date: "May 5, 2026"
date-modified: now
format:
  pdf:
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: right
    fig-align: center
    cap-location: top
    link-external-newwindow: true
    code-overflow: wrap
execute: 
  warning: false
---

## Introduction

### Background

Whether it be to celebrate a special occasion or relax after a long day, wine is one of the most popular and versatile beverages in the world. According to a 2019 survey of US adults, about three-quarters of Americans have a glass of wine each week, with about 44% of the US population having multiple glasses each week. Among those who reported drinking wine regularly, the most common reasons included liking the taste, viewing it as a stress reliever, enjoying it with their food, and believing that it makes them happier (Gitlin, 2019). 

One common source of wine production is Vinho Verde in northwest Portugal. The region contains about 17,269 hectares of vineyards and produces about 100 million liters of wine each year (“Vinho Verde: Wines Full of Atlantic Freshness”, 2026). The popularity of wine around the world as well as the major role of Vinho Verde in the global wine market necessitate a deeper understanding of wines produced from this region.

### Objective

Although most people have an allegiance to either red or white wine, few are familiar with the chemical composition of either type. Our goal is to analyze how several variables relating to wine chemistry differ between red and white wine. Furthermore, we aim to find how characteristics of wine samples from the Vinho Verde region correlate to their quality, and we will identify whether this relationship changes based on the wine’s color.

### The Dataset

All data in this report comes from a study presented by Cortez et al. in 2009. The period of data collection was from May of 2004 to February of 2007. 1,599 samples of red wine and 4,898 samples of white wine were analyzed, with the chemical characteristics of each being determined by the iLab software. Each wine sample was tasted and evaluated by three qualified judges who rated it on a scale of one to ten (with ten being the highest quality and one being the lowest quality). These findings were summarized into two csv files: one for red wine and one for white wine (Cortez et al., 2009).

## Wine Chemistry by Color

### Similar Characteristics

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

### Stark Contrasts

Text

## Correlates of Quality

### Strongest Predictors of Quality

Text

### Weakest Predictors of Quality

Text

### Varying Predictors of Quality by Color

Text

## Conclusion

### Takeaways

Text

### Next Steps

Text

## Code Appendix

## Conclusion

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
